---
title: "exploration"
output: html_document
date: "2024-11-20"
---

```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(ggrepel)
```

```{r}
# Load 2020 fires
beachie_2020 = rast("data/2020/beachie creek willamette 2020/BeachieCreek_SBS_final.tif")
freemont_2020 = rast("data/2020/freemont winema 2020/CRANE_sbs.tif")
holiday_2020 = rast("data/2020/holiday farm willamette 2020/HolidayFarm_SBS_final.tif")
hood_2020 = rast("data/2020/mt hood 2020/Riverside_SBS_final.tif")
umpqua_2020 = rast("data/2020/umpqua 2020/ArchieCreek_SBS_final.tif")

ben_young_2020 = rast("data/BEN_YOUNG_sbs/BEN_YOUNG_sbs.tif")
brattain_2020 = rast("data/BRATTAIN_sbs/BRATTAIN_sbs.tif")
lionshead_2020 = rast("data/lionshead_sbs/Lionshead_SBS_final.tif")
thielsen_2020 = rast("data/thielsen_sbs/Thielsen_SBS_final.tif")
two_four_2020 = rast("data/TWO_FOUR_TWO_sbs/TWO_FOUR_TWO_sbs.tif")
whiteriver_2020 = rast("data/whiteriver_sbs/WhiteRiver_SBS_final.tif")
```

```{r clean data}
# Adjust the scale of all fire rasters to be 1-4
adjust_scale <- function(raster) {
    raster[raster > 4] <- NA
}

# Align raster extents to a reference raster
align_extent <- function(raster, reference_raster) {
  extend(raster, ext(reference_raster))
}

adjust_scale(beachie_2020)
adjust_scale(freemont_2020)
adjust_scale(holiday_2020)
adjust_scale(hood_2020)
adjust_scale(umpqua_2020)
adjust_scale(ben_young_2020)
adjust_scale(brattain_2020)
adjust_scale(lionshead_2020)
adjust_scale(thielsen_2020)
adjust_scale(two_four_2020)
adjust_scale(whiteriver_2020)
```

```{r define functions}
# Reproject raster data function
reproject_raster <- function(raster, reference_raster) {
  project(raster, crs(reference_raster), res = res(reference_raster))
}

# Convert raster to dataframe function
raster_to_dataframe <- function(raster) {
  as.data.frame(raster, xy = TRUE)
}

# Assign nuneric values function for categorical variables for consistency
transform_soil_burn <- function(df) {
  # Check if 'SoilBurnSe' exists in the dataframe
  if ("SoilBurnSe" %in% colnames(df)) {
    df$SoilBurnSe <- as.character(df$SoilBurnSe)
    df$SoilBurnSe[df$SoilBurnSe == "High"] <- "4"
    df$SoilBurnSe[df$SoilBurnSe == "Moderate"] <- "3"
    df$SoilBurnSe[df$SoilBurnSe == "Low"] <- "2"
    df$SoilBurnSe[df$SoilBurnSe == "Unburned"] <- "1"
    df$SoilBurnSe <- as.factor(df$SoilBurnSe)
  } else{}
  return(df)
}

# Load and transform to a shapefile function
load_and_transform_shapefile <- function(shapefile_path, reference_raster) {
  shapefile <- st_read(shapefile_path)
  st_transform(shapefile, crs(reference_raster))
}

# Generate and save file function
plot_fires_and_bees <- function(willamette_df, rogue_df, umpqua_df, ecoregions, bee_sf, output_file) {
  ggplot() +
    geom_raster(data = willamette_df, aes(x = x, y = y, fill = SoilBurnSe)) +
    geom_raster(data = rogue_df, aes(x = x, y = y, fill = Layer_1)) +
    geom_raster(data = umpqua_df, aes(x = x, y = y, fill = Layer_1)) + 
    geom_sf(data = ecoregions, fill = NA, color = "black") + 
    geom_sf(data = bee_sf, aes(color = Collection.Month), size = 3) +
    scale_fill_brewer(palette = "Spectral", direction = -1) +
    scale_color_viridis_d(name = "Collection Month & Year") +
    labs(
      title = "Map of Longula Bee Species Locations by Collection Month and Year with 3 2018 Fires Plotted",
      x = "Longitude",
      y = "Latitude"
    ) +
    annotation_scale(location = "br", width_hint = 0.2)
  
  ggsave(output_file, width = 8, height = 6, dpi = 300)
}

# Function to process a list of rasters
process_rasters <- function(raster, reference_raster) {
  # Reproject the raster
  raster <- reproject_raster(raster, reference_raster)
    
  # Convert raster to data frame
  raster_df <- raster_to_dataframe(raster)

  return(raster_df)
}
```

```{r clean data 2}
# Apply the function for each fire
beachie_2020_df <- process_rasters(beachie_2020, beachie_2020)
freemont_2020_df <- process_rasters(freemont_2020, beachie_2020)
holiday_2020_df <- process_rasters(holiday_2020, beachie_2020)
hood_2020_df <- process_rasters(hood_2020, beachie_2020)
umpqua_2020_df <- process_rasters(umpqua_2020, beachie_2020)

ben_young_2020_df <- process_rasters(ben_young_2020, beachie_2020)
brattain_2020_df <- process_rasters(brattain_2020, beachie_2020)
lionshead_2020_df <- process_rasters(lionshead_2020, beachie_2020)
thielsen_2020_df <- process_rasters(thielsen_2020, beachie_2020)
two_four_2020_df <- process_rasters(two_four_2020, beachie_2020)
whiteriver_2020_df <- process_rasters(whiteriver_2020, beachie_2020)

# Convert the columns to numeric (continuous) if they aren't already
beachie_2020_df$Layer_1 <- as.numeric(beachie_2020_df$Layer_1)
freemont_2020_df$Layer_1 <- as.numeric(freemont_2020_df$Layer_1)
holiday_2020_df$Layer_1 <- as.numeric(holiday_2020_df$Layer_1)
hood_2020_df$Layer_1 <- as.numeric(hood_2020_df$Layer_1)
umpqua_2020_df$Layer_1 <- as.numeric(umpqua_2020_df$Layer_1)

ben_young_2020_df$Layer_1 <- as.numeric(ben_young_2020_df$Layer_1)
brattain_2020_df$Layer_1 <- as.numeric(brattain_2020_df$Layer_1)
lionshead_2020_df$Layer_1 <- as.numeric(lionshead_2020_df$Layer_1)
thielsen_2020_df$Layer_1 <- as.numeric(thielsen_2020_df$Layer_1)
two_four_2020_df$Layer_1 <- as.numeric(two_four_2020_df$Layer_1)
whiteriver_2020_df$Layer_1 <- as.numeric(whiteriver_2020_df$Layer_1)
```

```{r load oregon data}
oregon_ecoregions <- st_read("data/OR-ecoregions/Ecoregions_OregonConservationStrategy.shp")
oregon_ecoregions <- st_transform(oregon_ecoregions, crs(beachie_2020))
```

```{r plot fires}
ggplot() +
  geom_raster(data = beachie_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = freemont_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = holiday_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = hood_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = umpqua_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = ben_young_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = brattain_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = lionshead_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = thielsen_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = two_four_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = whiteriver_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  
  # Add the Oregon ecoregions as a boundary layer
  geom_sf(data = oregon_ecoregions, fill = NA, color = "black") +  
  
  # Customize colors
  scale_fill_gradient(low = "blue", high = "red", name = "Intensity") +  # Set legend title to "Intensity"

  # Add labels and scale
  labs(
    title = "Map of Fire Impacts and Bee Species Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # Add annotation for scale
  annotation_scale(location = "br", width_hint = 0.2) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10))
```

### Initial look at the Oregon Bee Atlas data
First, let's read the OBA data and see what columns we have
```{r}
oba_df <- read.csv("data/OBA_2018-2023.csv")
head(oba_df)
```
Now, let's plot all of the bees that are within Oregon on top of a map to see potential areas of interest.
```{r}
# Step 1: Retrieve Oregon county polygon data
or <- map_data("county", "oregon") %>%  
  select(lon = long, lat, group, id = subregion)

# Convert the Oregon polygon to an sf object (simple feature)
or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326)

# Step 2: Convert your bee data (`oba_df`) to an sf object
## Also going to normalize Genus and Species values
bees_sf <- oba_df %>%
  select(Dec..Lat., Dec..Long., Genus, Species) %>%
  rename(Latitude = Dec..Lat., Longitude = Dec..Long.) %>%
  mutate(
    Genus = tolower(trimws(Genus)), Species = tolower(trimws(Species))
  ) %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)


# Step 3: Crop the bee data to the Oregon bounds
bees_in_oregon <- st_crop(bees_sf, or_sf)

# Step 4: Plot the polygon (oregon), then plot the spatial fetaures (bee locations) on top
ggplot() +
  # Plot Oregon polygons
  geom_polygon(data = or, aes(x = lon, y = lat, group = group), 
               fill = "lightgreen", color = "darkgreen") +
  # Overlay bee locations with coloring by genus
  geom_sf(data = bees_in_oregon) +
  labs(title = "Bees in Oregon", x = "Longitude", y = "Latitude") +
  theme(
    legend.key.size = unit(0.4, "cm"), # Reduce key size
    legend.text = element_text(size = 4), # Reduce text size
    legend.title = element_text(size = 8) # Reduce title size
  ) +
  guides(color = guide_legend(ncol = 3))
```

### Plot the fires seperately
We want to:
- Plot the bees in the general region the year before the fire
- Then plot the bees in the general region of the fire after the fire year
- Going to look at fires from just 2020

MonthAb contains the month as a number, as expected
Year.1 contains the year as a YYYY number, as expected
```{r}
# Look at the columns of the bee stuff first
oba_df <- oba_df %>%
  rename(
    year = Year.1,
    month = MonthAb,
    full_date = Collection.Date,
    latitude = Dec..Lat.,
    longitude = Dec..Long.,
    plant = Associated.plant
  )

oba_selection <- oba_df %>%
  select(year, month, Genus, Species, plant, latitude, longitude) %>%
  filter(!is.na(Genus), Genus != "")

oba_selection
```


```{r splitting data into before and after fires}
bees_before_fire <- function(fire_month) {
  oba_selection %>%
    filter(
      #(year %in% c(2019)) |  # Include all data for 2018 and 2019
      (year == 2020 & month %in% 1:fire_month))}  # Include only up to the specified month in 2020

bees_after_fire <- function(fire_month) {
  oba_selection %>%
    filter(
      #(year %in% c(2021)) |  # Include all data for 2021 and 2022
      (year == 2020 & month %in% fire_month:12))}  # Include only the specified month and after in 2020
```

```{r}  
plot_bees_in_fire_region <- function(bee_data, fire_sp_obj, fire_name) {
  ### This function will take in bee data (as a dataframe) and a fire raster
  ### and plot the bees found within a 50 km radius of the provided fire
  ### it will also provide a count of the bees within the radius
  ### The goal is to be able to compare the bees in the radius before and after the fire
  
  
  ## Deal with the base map of Oregon
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  
  ## Make sure fire spatial obj matches Oregon map's CRS
  fire_sp_obj <- project(fire_sp_obj, crs(or_sf))
  
  ## Convert bee data to an sf object & make sure CRS matches
  bees_sf <- bee_data %>%
    mutate(
      Genus = tolower(trimws(Genus)),
      Species = tolower(trimws(Species))
    ) %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  
  bees_sf <- st_transform(bees_sf, crs = st_crs(or_sf))
  
  ## Use the fire's bounds to create a buffer (25 km) & capture all the bees within it by cropping
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 15000)
  
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)
  
  ## Total number of bees in the buffer for comparison (before / after fire)
  total_bees <- nrow(bees_in_fire_region)
  
  ## Plot the Oregon polygon and the bee locations in the fire region
  ggplot() +
    geom_polygon(data = or, aes(x = lon, y = lat, group = group),
                 fill = "lightgreen", color = "darkgreen") +
    
    geom_sf(data = fire_buffer, fill = NA, color = "red", linetype = "dashed") +
    
    geom_sf(data = bees_in_fire_region, size = 2, alpha = 0.7) +
    
    labs(
      title = paste("Bees in Fire Region:", fire_name),
      subtitle = paste("Total bees found in the buffer region:", total_bees),
      x = "Longitude", 
      y = "Latitude"
    )
}
```

```{r}
# Test the function with different fire spatial objects
plot_bees_in_fire_region(bees_before_fire(9), umpqua_2020, "Umpqua Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(9), hood_2020, "Hood Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(9), holiday_2020, "Holiday Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(8), beachie_2020, "Beachie Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(8), freemont_2020, "Freemont Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(7), ben_young_2020, "Ben Young Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(9), brattain_2020, "Brattain Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(9), thielsen_2020, "Thielsen Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(9), two_four_2020, "Two Four Fire (before 2020)")
plot_bees_in_fire_region(bees_before_fire(8), whiteriver_2020, "Whiteriver Fire (before 2020)")

plot_bees_in_fire_region(bees_after_fire(9), umpqua_2020, "Umpqua Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(9), hood_2020, "Hood Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(9), holiday_2020, "Holiday Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(8), beachie_2020, "Beachie Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(8), freemont_2020, "Freemont Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(7), ben_young_2020, "Ben Young Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(9), brattain_2020, "Brattain Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(9), thielsen_2020, "Thielsen Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(9), two_four_2020, "Two Four Fire (after 2020)")
plot_bees_in_fire_region(bees_after_fire(8), whiteriver_2020, "Whiteriver Fire (after 2020)")
```
### Define the "average severity" for a given fire

```{r average severity}
define_avg_severity <- function(fire_df) {
  ### This function takes all the severity data ("Layer_1" in the raster)
  ### returns the average severity for the fire
  most_severe <- fire_df %>% filter(Layer_1 == 5 | Layer_1 == 4)
  avg_severity <- sum(most_severe$Layer_1) / nrow(fire_df)
  return(avg_severity)
}

define_avg_severity(umpqua_2020_df)
define_avg_severity(hood_2020_df)
define_avg_severity(holiday_2020_df)
define_avg_severity(beachie_2020_df)
define_avg_severity(freemont_2020_df)
define_avg_severity(ben_young_2020_df)
define_avg_severity(brattain_2020_df)
define_avg_severity(lionshead_2020_df)
define_avg_severity(thielsen_2020_df)
define_avg_severity(two_four_2020_df)
define_avg_severity(whiteriver_2020_df)
```
## Statistical test to evaluate significance in change

```{r}
bees_in_buffer_zone <- function(bee_data, fire_sp_obj) {
  
  ## Deal with the base map of Oregon
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  
  ## Make sure fire spatial obj matches Oregon map's CRS
  fire_sp_obj <- project(fire_sp_obj, crs(or_sf))
  
  ## Convert bee data to an sf object & make sure CRS matches
  bees_sf <- bee_data %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  
  bees_sf <- st_transform(bees_sf, crs = st_crs(or_sf))
  
  ## Use the fire's bounds to create a buffer (50 km) & capture all the bees within it by cropping
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 25000)
  
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)

  ## Total number of bees in the buffer for comparison
  total_bees <- nrow(bees_in_fire_region)
  return(total_bees)
}
```

```{r}
# Calculate bees in buffer zone before and after the fire for Umpqua fire
percent_change <- function(fire_rast, bees_before, bees_after){
  before_fire <- bees_in_buffer_zone(bees_before, fire_rast)
  after_fire <- bees_in_buffer_zone(bees_after, fire_rast)
  umpqua_perc_change <- calculate_percent_change(before_fire, after_fire)
  }

umpqua_perc_change <- percent_change(umpqua_2020, bees_before_fire(9), bees_after_fire(9))
hood_perc_change <- percent_change(hood_2020, bees_before_fire(9), bees_after_fire(9))
holiday_perc_change <- percent_change(holiday_2020, bees_before_fire(9), bees_after_fire(9))
beachie_perc_change <- percent_change(beachie_2020, bees_before_fire(8), bees_after_fire(8))
freemont_perc_change <- percent_change(freemont_2020, bees_before_fire(8), bees_after_fire(8))

ben_young_perc_change <- percent_change(ben_young_2020, bees_before_fire(7), bees_after_fire(7))
brattain_perc_change <- percent_change(brattain_2020, bees_before_fire(9), bees_after_fire(9))
lionshead_perc_change <- percent_change(lionshead_2020, bees_before_fire(9), bees_after_fire(9))
thielsen_perc_change <- percent_change(thielsen_2020, bees_before_fire(9), bees_after_fire(9))
two_four_perc_change <- percent_change(two_four_2020, bees_before_fire(9), bees_after_fire(9))
whiteriver_perc_change <- percent_change(whiteriver_2020, bees_before_fire(8), bees_after_fire(8))
```

```{r}
print("Umpqua average severity and percent change in bees")
define_avg_severity(umpqua_2020_df)
umpqua_perc_change

print("Hood average severity and percent change in bees")
define_avg_severity(hood_2020_df)
hood_perc_change

print("Holiday average severity and percent change in bees")
define_avg_severity(holiday_2020_df)
holiday_perc_change

print("Beachie average severity and percent change in bees")
define_avg_severity(beachie_2020_df)
beachie_perc_change

print("Freemont average severity and percent change in bees")
define_avg_severity(freemont_2020_df)
freemont_perc_change

print("Ben Young average severity and percent change in bees")
define_avg_severity(ben_young_2020_df)
ben_young_perc_change

print("Brattain average severity and percent change in bees")
define_avg_severity(brattain_2020_df)
brattain_perc_change

print("Lionshead average severity and percent change in bees")
define_avg_severity(lionshead_2020_df)
lionshead_perc_change

print("Thielsen average severity and percent change in bees")
define_avg_severity(thielsen_2020_df)
thielsen_perc_change

print("Two Four average severity and percent change in bees")
define_avg_severity(two_four_2020_df)
two_four_perc_change

print("Whiteriver average severity and percent change in bees")
define_avg_severity(whiteriver_2020_df)
whiteriver_perc_change
```

```{r}
# Create a data frame with average severities and percent changes for all fires
fires_data <- data.frame(
  fire = c("Umpqua", "Hood", "Holiday", "Beachie", "Freemont", 
           "Ben Young", "Brattain", "Lionshead", "Thielsen", "Two Four"),
  avg_severity = c(define_avg_severity(umpqua_2020_df),
                   define_avg_severity(hood_2020_df),
                   define_avg_severity(holiday_2020_df),
                   define_avg_severity(beachie_2020_df),
                   define_avg_severity(freemont_2020_df),
                   define_avg_severity(ben_young_2020_df),
                   define_avg_severity(brattain_2020_df),
                   define_avg_severity(lionshead_2020_df),
                   define_avg_severity(thielsen_2020_df),
                   define_avg_severity(two_four_2020_df)),
  percent_change = c(umpqua_perc_change,
                     hood_perc_change,
                     holiday_perc_change,
                     beachie_perc_change,
                     freemont_perc_change,
                     ben_young_perc_change,
                     brattain_perc_change,
                     lionshead_perc_change,
                     thielsen_perc_change,
                     two_four_perc_change)
)
```

```{r}
# Correlation Analysis
correlation <- cor(fires_data$avg_severity, fires_data$percent_change, method = "pearson")
cat("Pearson correlation coefficient:", correlation, "\n")

# Regression Plot
ggplot(fires_data, aes(x = avg_severity, y = percent_change, label = fire)) +
  geom_point(size = 4, color = "#1f78b4", alpha = 0.8) +
  geom_smooth(method = "lm", color = "#e31a1c", se = TRUE) +
  geom_text_repel(
    size = 5, color = "black", box.padding = 0.35, max.overlaps = 10,
    segment.size = 0.5, segment.color = "gray"
  ) +
  geom_hline(
    yintercept = 0, 
    size = 0.25
  ) + 
  labs(
    title = paste("Percent Change in Bees vs. Average Fire Severity, r =", round(correlation, 2)),
    x = "Average Fire Severity",
    y = "Percent Change in Bee Counts",
    caption = "Data source: Oregon Bee Atlas and BAER"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 12),
    plot.caption = element_text(size = 10, hjust = 1, face = "italic")
  )
```

```{r} 
```