---
title: "exploration"
output: html_document
date: "2024-11-20"
---

```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(ggrepel)
```

```{r}
# Load 2020 fires
beachie_2020 = rast("data/2020/beachie creek willamette 2020/BeachieCreek_SBS_final.tif")
freemont_2020 = rast("data/2020/freemont winema 2020/CRANE_sbs.tif")
holiday_2020 = rast("data/2020/holiday farm willamette 2020/HolidayFarm_SBS_final.tif")
hood_2020 = rast("data/2020/mt hood 2020/Riverside_SBS_final.tif")
umpqua_2020 = rast("data/2020/umpqua 2020/ArchieCreek_SBS_final.tif")

ben_young_2020 = rast("data/BEN_YOUNG_sbs/BEN_YOUNG_sbs.tif")
brattain_2020 = rast("data/BRATTAIN_sbs/BRATTAIN_sbs.tif")
lionshead_2020 = rast("data/lionshead_sbs/Lionshead_SBS_final.tif")
thielsen_2020 = rast("data/thielsen_sbs/Thielsen_SBS_final.tif")
two_four_2020 = rast("data/TWO_FOUR_TWO_sbs/TWO_FOUR_TWO_sbs.tif")
whiteriver_2020 = rast("data/whiteriver_sbs/WhiteRiver_SBS_final.tif")
```

```{r clean data}
# Adjust the scale of all fire rasters to be 1-4
adjust_scale <- function(raster) {
    raster[raster > 4] <- NA
    return(raster)
}

# Align raster extents to a reference raster
align_extent <- function(raster, reference_raster) {
  return(extend(raster, ext(reference_raster)))
}

# Apply the function and reassign
beachie_2020 <- adjust_scale(beachie_2020)
freemont_2020 <- adjust_scale(freemont_2020)
holiday_2020 <- adjust_scale(holiday_2020)
hood_2020 <- adjust_scale(hood_2020)
umpqua_2020 <- adjust_scale(umpqua_2020)
ben_young_2020 <- adjust_scale(ben_young_2020)
brattain_2020 <- adjust_scale(brattain_2020)
lionshead_2020 <- adjust_scale(lionshead_2020)
thielsen_2020 <- adjust_scale(thielsen_2020)
two_four_2020 <- adjust_scale(two_four_2020)
whiteriver_2020 <- adjust_scale(whiteriver_2020)
```

```{r define functions}
# Reproject raster data function
reproject_raster <- function(raster, reference_raster) {
  return(project(raster, crs(reference_raster), res = res(reference_raster)))
}

# Convert raster to dataframe function
raster_to_dataframe <- function(raster) {
  return(as.data.frame(raster, xy = TRUE))
}

# Assign nuneric values function for categorical variables for consistency
transform_soil_burn <- function(df) {
  # Check if 'SoilBurnSe' exists in the dataframe
  if ("SoilBurnSe" %in% colnames(df)) {
    df$SoilBurnSe <- as.character(df$SoilBurnSe)
    df$SoilBurnSe[df$SoilBurnSe == "High"] <- "4"
    df$SoilBurnSe[df$SoilBurnSe == "Moderate"] <- "3"
    df$SoilBurnSe[df$SoilBurnSe == "Low"] <- "2"
    df$SoilBurnSe[df$SoilBurnSe == "Unburned"] <- "1"
    df$SoilBurnSe <- as.factor(df$SoilBurnSe)
  } else{}
  return(df)
}

# Load and transform to a shapefile function
load_and_transform_shapefile <- function(shapefile_path, reference_raster) {
  shapefile <- st_read(shapefile_path)
  return(st_transform(shapefile, crs(reference_raster)))
}

# Function to process a list of rasters
process_rasters <- function(raster, reference_raster) {
  # Reproject the raster
  raster <- reproject_raster(raster, reference_raster)
    
  # Convert raster to data frame
  raster_df <- raster_to_dataframe(raster)

  return(raster_df)
}
```

```{r clean data 2}
# Apply the function for each fire
beachie_2020_df <- process_rasters(beachie_2020, beachie_2020)
freemont_2020_df <- process_rasters(freemont_2020, beachie_2020)
holiday_2020_df <- process_rasters(holiday_2020, beachie_2020)
hood_2020_df <- process_rasters(hood_2020, beachie_2020)
umpqua_2020_df <- process_rasters(umpqua_2020, beachie_2020)

ben_young_2020_df <- process_rasters(ben_young_2020, beachie_2020)
brattain_2020_df <- process_rasters(brattain_2020, beachie_2020)
lionshead_2020_df <- process_rasters(lionshead_2020, beachie_2020)
thielsen_2020_df <- process_rasters(thielsen_2020, beachie_2020)
two_four_2020_df <- process_rasters(two_four_2020, beachie_2020)
whiteriver_2020_df <- process_rasters(whiteriver_2020, beachie_2020)

# Convert the columns to numeric (continuous) if they aren't already
beachie_2020_df$Layer_1 <- as.numeric(beachie_2020_df$Layer_1)
freemont_2020_df$Layer_1 <- as.numeric(freemont_2020_df$Layer_1)
holiday_2020_df$Layer_1 <- as.numeric(holiday_2020_df$Layer_1)
hood_2020_df$Layer_1 <- as.numeric(hood_2020_df$Layer_1)
umpqua_2020_df$Layer_1 <- as.numeric(umpqua_2020_df$Layer_1)

ben_young_2020_df$Layer_1 <- as.numeric(ben_young_2020_df$Layer_1)
brattain_2020_df$Layer_1 <- as.numeric(brattain_2020_df$Layer_1)
lionshead_2020_df$Layer_1 <- as.numeric(lionshead_2020_df$Layer_1)
thielsen_2020_df$Layer_1 <- as.numeric(thielsen_2020_df$Layer_1)
two_four_2020_df$Layer_1 <- as.numeric(two_four_2020_df$Layer_1)
whiteriver_2020_df$Layer_1 <- as.numeric(whiteriver_2020_df$Layer_1)
```

```{r load oregon data}
oregon_ecoregions <- st_read("data/OR-ecoregions/Ecoregions_OregonConservationStrategy.shp")
oregon_ecoregions <- st_transform(oregon_ecoregions, crs(beachie_2020))
```

```{r plot fires}
ggplot() +
  geom_raster(data = beachie_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = freemont_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = holiday_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = hood_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = umpqua_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = ben_young_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = brattain_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = lionshead_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = thielsen_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = two_four_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = whiteriver_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  
  # Add the Oregon ecoregions as a boundary layer
  geom_sf(data = oregon_ecoregions, fill = NA, color = "black") +  
  
  # Customize colors
  scale_fill_gradient(low = "blue", high = "red", name = "Intensity") + 

  # Add labels and scale
  labs(
    title = "Map of Fire Impacts and Bee Species Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # Add annotation for scale
  annotation_scale(location = "br", width_hint = 0.2) +
  theme_minimal() +
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10))
```

### Initial look at the Oregon Bee Atlas data
First, let's read the OBA data and see what columns we have
```{r}
oba_df <- read.csv("data/OBA_2018-2023.csv")
head(oba_df)
```
Now, let's plot all of the bees that are within Oregon on top of a map to see potential areas of interest.
```{r}
# Step 1: Retrieve Oregon county polygon data
or <- map_data("county", "oregon") %>%  
  select(lon = long, lat, group, id = subregion)

# Convert the Oregon polygon to an sf object (simple feature)
or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326)

# Step 2: Convert your bee data (`oba_df`) to an sf object
## Also going to normalize Genus and Species values
bees_sf <- oba_df %>%
  select(Dec..Lat., Dec..Long., Genus, Species) %>%
  rename(Latitude = Dec..Lat., Longitude = Dec..Long.) %>%
  mutate(
    Genus = tolower(trimws(Genus)), Species = tolower(trimws(Species))
  ) %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)


# Step 3: Crop the bee data to the Oregon bounds
bees_in_oregon <- st_crop(bees_sf, or_sf)

# Step 4: Plot the polygon (oregon), then plot the spatial fetaures (bee locations) on top
ggplot() +
  # Plot Oregon polygons
  geom_polygon(data = or, aes(x = lon, y = lat, group = group), 
               fill = "lightgreen", color = "darkgreen") +
  # Overlay bee locations with coloring by genus
  geom_sf(data = bees_in_oregon) +
  labs(title = "Bees in Oregon", x = "Longitude", y = "Latitude") +
  theme(
    legend.key.size = unit(0.4, "cm"), # Reduce key size
    legend.text = element_text(size = 4), # Reduce text size
    legend.title = element_text(size = 8) # Reduce title size
  ) +
  guides(color = guide_legend(ncol = 3))
```

### Plot the fires seperately
We want to:
- Plot the bees in the general region the year before the fire
- Then plot the bees in the general region of the fire after the fire year
- Going to look at fires from just 2020

MonthAb contains the month as a number, as expected
Year.1 contains the year as a YYYY number, as expected
```{r}
# Look at the columns of the bee stuff first
oba_df <- oba_df %>%
  rename(
    year = Year.1,
    month = MonthAb,
    full_date = Collection.Date,
    latitude = Dec..Lat.,
    longitude = Dec..Long.,
    plant = Associated.plant
  )

oba_selection <- oba_df %>%
  select(year, month, Genus, Species, plant, latitude, longitude) %>%
  filter(!is.na(Genus), Genus != "")

oba_selection
```


```{r splitting data into before and after fires}
# Function to get bee data 24 months before the fire for a fire in 2020 that occured in fire_month
bees_before_fire <- function(fire_month) {
  oba_selection %>%
    filter(
      # Ex: fire in Sept 2020 would want to get data from Sept 2018 up to Sept 2020
      (year == 2020 & month < fire_month) |  # Bees in Jan - Oct 2020
      (year == 2019) |                       # Bees in all of 2019
      (year == 2018 & month >= fire_month)   # Bees in Sept - Dec 2018
    )
}

# Function to get bee data 24 months after the fire for a fire in 2020 that occured in fire_month
bees_after_fire <- function(fire_month) {
  oba_selection %>%
    filter(
      # Ex: fire in Sept 2020 would want to get data from Sept 2020 up to Sept 2022
      (year == 2020 & month > fire_month) |   # Bees in Oct - Dec 2020
      (year == 2021) |                        # Bees in all of 2021
      (year == 2022 & month <= fire_month)    # Bees in Jan - Sept 2022
    )
}
before_sept_20 <- bees_before_fire(9)
before_sept_20
```

```{r}  
plot_bees_in_fire_region <- function(bee_data, fire_sp_obj, fire_name) {
  ### This function will take in bee data (as a dataframe) and a fire raster
  ### and plot the bees found within a 50 km radius of the provided fire
  ### it will also provide a count of the bees within the radius
  ### The goal is to be able to compare the bees in the radius before and after the fire
  
  
  ## Deal with the base map of Oregon
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  
  ## Make sure fire spatial obj matches Oregon map's CRS
  fire_sp_obj <- project(fire_sp_obj, crs(or_sf))
  
  ## Convert bee data to an sf object & make sure CRS matches
  bees_sf <- bee_data %>%
    mutate(
      Genus = tolower(trimws(Genus)),
      Species = tolower(trimws(Species))
    ) %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  

  bees_sf <- st_transform(bees_sf, crs = st_crs(or_sf))
  
  ## Use the fire's bounds to create a buffer (25 km) & capture all the bees within it by cropping
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 25000)
  
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)
  
  ## Total number of bees in the buffer for comparison (before / after fire)
  total_bees <- nrow(bees_in_fire_region)
  
  ## Plot the Oregon polygon and the bee locations in the fire region
  ggplot() +
    geom_polygon(data = or, aes(x = lon, y = lat, group = group),
                 fill = "lightgreen", color = "darkgreen") +
    
    geom_sf(data = fire_buffer, fill = NA, color = "red", linetype = "dashed") +
    
    geom_sf(data = bees_in_fire_region, size = 2, alpha = 0.7) +
    
    labs(
      title = paste("Bees in Fire Region:", fire_name),
      subtitle = paste("Total bees found in the buffer region:", total_bees),
      x = "Longitude", 
      y = "Latitude"
    )
}
```

```{r}
# Test the function with different fire spatial objects
plot_bees_in_fire_region(bees_before_fire(9), umpqua_2020, "Umpqua Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), umpqua_2020, "Umpqua Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(9), hood_2020, "Hood Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), hood_2020, "Hood Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(9), holiday_2020, "Holiday Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), holiday_2020, "Holiday Fire (2 years after fire))")

plot_bees_in_fire_region(bees_before_fire(8), beachie_2020, "Beachie Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(8), beachie_2020, "Beachie Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(8), freemont_2020, "Freemont Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(8), freemont_2020, "Freemont Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(7), ben_young_2020, "Ben Young Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(7), ben_young_2020, "Ben Young Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(8), lionshead_2020, "Lionshead Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(8), lionshead_2020, "Lionshead Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(9), brattain_2020, "Brattain Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), brattain_2020, "Brattain Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(9), thielsen_2020, "Thielsen Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), thielsen_2020, "Thielsen Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(9), two_four_2020, "Two Four Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(9), two_four_2020, "Two Four Fire (2 years after fire)")

plot_bees_in_fire_region(bees_before_fire(8), whiteriver_2020, "Whiteriver Fire (2 years before fire)")
plot_bees_in_fire_region(bees_after_fire(8), whiteriver_2020, "Whiteriver Fire (2 years after fire)")
```
### Define the "average severity" for a given fire

```{r average severity}
define_avg_severity <- function(fire_df) {
  ### This function takes all the severity data ("Layer_1" in the raster)
  ### returns the average severity for the fire
  prop_severe <- sum(fire_df$Layer_1 == 4 | fire_df$Layer_1 == 5) / nrow(fire_df)
  avg_severity <- sum(fire_df$Layer_1) / nrow(fire_df)
  median_severity <- median(fire_df$Layer_1)
  mode_severity <- as.numeric(names(sort(table(fire_df$Layer_1), decreasing = TRUE)[1]))
  return(avg_severity)
}

define_avg_severity(umpqua_2020_df)
define_avg_severity(hood_2020_df)
define_avg_severity(holiday_2020_df)
define_avg_severity(beachie_2020_df)
define_avg_severity(freemont_2020_df)
define_avg_severity(ben_young_2020_df)
define_avg_severity(brattain_2020_df)
define_avg_severity(lionshead_2020_df)
define_avg_severity(thielsen_2020_df)
define_avg_severity(two_four_2020_df)
define_avg_severity(whiteriver_2020_df)
```
## Statistical test to evaluate significance in change

```{r}
bees_in_buffer_zone <- function(bee_data, fire_sp_obj) {
  
  ## Deal with the base map of Oregon
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  
  ## Make sure fire spatial obj matches Oregon map's CRS
  fire_sp_obj <- project(fire_sp_obj, crs(or_sf))
  
  ## Convert bee data to an sf object & make sure CRS matches
  bees_sf <- bee_data %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  
  bees_sf <- st_transform(bees_sf, crs = st_crs(or_sf))
  
  ## Use the fire's bounds to create a buffer (50 km) & capture all the bees within it by cropping
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 25000)
  
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)

  ## Total number of bees in the buffer for comparison
  total_bees <- nrow(bees_in_fire_region)
  return(total_bees)
}
```

```{r}
# Calculate bees in buffer zone before and after the fire for Umpqua fire
percent_change <- function(fire_rast, bees_before, bees_after) {
  before_fire <- bees_in_buffer_zone(bees_before, fire_rast)
  after_fire <- bees_in_buffer_zone(bees_after, fire_rast)
  percent_change_value <- ((after_fire - before_fire) / before_fire) * 100
  return(percent_change_value)}

umpqua_perc_change <- percent_change(umpqua_2020, bees_before_fire(9), bees_after_fire(9))
hood_perc_change <- percent_change(hood_2020, bees_before_fire(9), bees_after_fire(9))
holiday_perc_change <- percent_change(holiday_2020, bees_before_fire(9), bees_after_fire(9))
beachie_perc_change <- percent_change(beachie_2020, bees_before_fire(8), bees_after_fire(8))
freemont_perc_change <- percent_change(freemont_2020, bees_before_fire(8), bees_after_fire(8))

ben_young_perc_change <- percent_change(ben_young_2020, bees_before_fire(7), bees_after_fire(7))
brattain_perc_change <- percent_change(brattain_2020, bees_before_fire(9), bees_after_fire(9))
lionshead_perc_change <- percent_change(lionshead_2020, bees_before_fire(8), bees_after_fire(9))
thielsen_perc_change <- percent_change(thielsen_2020, bees_before_fire(9), bees_after_fire(9))
two_four_perc_change <- percent_change(two_four_2020, bees_before_fire(9), bees_after_fire(9))
whiteriver_perc_change <- percent_change(whiteriver_2020, bees_before_fire(8), bees_after_fire(8))
```

```{r}
# Create an empty data frame to store the results
severity_data <- data.frame(
  Location = character(),
  Average_Severity = numeric(),
  Percent_Change_in_Bees = numeric(),
  stringsAsFactors = FALSE
)

# Function to add data to the data frame
add_to_dataframe <- function(location, avg_severity, perc_change) {
  severity_data <<- rbind(severity_data, data.frame(
    Location = location,
    Average_Severity = avg_severity,
    Percent_Change_in_Bees = perc_change
  ))
}

# Example code using your original function
# Umpqua
avg_severity <- define_avg_severity(umpqua_2020_df)
add_to_dataframe("Umpqua", avg_severity, umpqua_perc_change)

# Hood
avg_severity <- define_avg_severity(hood_2020_df)
add_to_dataframe("Hood", avg_severity, hood_perc_change)

# Holiday
avg_severity <- define_avg_severity(holiday_2020_df)
add_to_dataframe("Holiday", avg_severity, holiday_perc_change)

# Beachie
avg_severity <- define_avg_severity(beachie_2020_df)
add_to_dataframe("Beachie", avg_severity, beachie_perc_change)

# Freemont
avg_severity <- define_avg_severity(freemont_2020_df)
add_to_dataframe("Freemont", avg_severity, freemont_perc_change)

# Ben Young
avg_severity <- define_avg_severity(ben_young_2020_df)
add_to_dataframe("Ben Young", avg_severity, ben_young_perc_change)

# Brattain
avg_severity <- define_avg_severity(brattain_2020_df)
add_to_dataframe("Brattain", avg_severity, brattain_perc_change)

# Lionshead
avg_severity <- define_avg_severity(lionshead_2020_df)
add_to_dataframe("Lionshead", avg_severity, lionshead_perc_change)

# Thielsen
avg_severity <- define_avg_severity(thielsen_2020_df)
add_to_dataframe("Thielsen", avg_severity, thielsen_perc_change)

# Two Four
avg_severity <- define_avg_severity(two_four_2020_df)
add_to_dataframe("Two Four", avg_severity, two_four_perc_change)

# Whiteriver
avg_severity <- define_avg_severity(whiteriver_2020_df)
add_to_dataframe("Whiteriver", avg_severity, whiteriver_perc_change)

# View the final data frame
severity_data
```

```{r}
# Create a data frame with average severities and percent changes for all fires
fires_data <- data.frame(
  fire = c("Umpqua", "Hood", "Holiday", "Beachie", "Lionshead", "Thielsen", "Freemont", "Ben Young", "Brattain", "Two Four", "Whiteriver"),
  avg_severity = c(define_avg_severity(umpqua_2020_df),
                   define_avg_severity(hood_2020_df),
                   define_avg_severity(holiday_2020_df),
                   define_avg_severity(beachie_2020_df),
                   define_avg_severity(lionshead_2020_df),
                   define_avg_severity(thielsen_2020_df),
                   define_avg_severity(freemont_2020_df), 
                   define_avg_severity(ben_young_2020_df), 
                   define_avg_severity(brattain_2020_df), 
                   define_avg_severity(two_four_2020_df),
                   define_avg_severity(whiteriver_2020_df)),
  percent_change = c(umpqua_perc_change,
                     hood_perc_change,
                     holiday_perc_change,
                     beachie_perc_change,
                     lionshead_perc_change,
                     thielsen_perc_change,
                     freemont_perc_change, 
                     ben_young_perc_change, 
                     brattain_perc_change, 
                     two_four_perc_change,
                     whiteriver_perc_change))
```

```{r}
# Correlation Analysis
correlation <- cor(fires_data$avg_severity, fires_data$percent_change, method = "pearson")
cat("Pearson correlation coefficient:", correlation, "\n")

# Regression Plot
ggplot(fires_data, aes(x = avg_severity, y = percent_change, label = fire)) +
  geom_point(size = 4, color = "#1f78b4", alpha = 0.8) +
  geom_smooth(method = "lm", color = "#e31a1c", se = TRUE) +
  geom_text_repel(
    size = 5, color = "black", box.padding = 0.35, max.overlaps = 10,
    segment.size = 0.5, segment.color = "gray"
  ) +
  geom_hline(
    yintercept = 0, 
    size = 0.25
  ) + 
  labs(
    title = paste("Percent Change in Bees vs. Average Fire Severity, r =", round(correlation, 2)),
    x = "Average Fire Severity",
    y = "Percent Change in Bee Counts",
    caption = "Data source: Oregon Bee Atlas and BAER"
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "italic"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 12),
    plot.caption = element_text(size = 10, hjust = 1, face = "italic")
  )
```
Null Hypothesis: There is no linear relationship between fire soil burn severity and % change in bee count. r = 0
Alternative Hypothesis: There is a linear relationship between fire soil burn severity and % change in bee count. r != 0
```{r} 
# Perform the Pearson correlation test
cor_test <- cor.test(fires_data$avg_severity, fires_data$percent_change, method = "pearson")

# Print the results
cat("Pearson correlation coefficient:", cor_test$estimate, "\n")
cat("P-value:", cor_test$p.value, "\n")

# Check significance level
if (cor_test$p.value < 0.05) {
  cat("Result: Reject the null hypothesis. There is evidence of a linear relationship.\n")
} else {
  cat("Result: Fail to reject the null hypothesis. No evidence of a linear relationship.\n")
}
```


Null Hypothesis: There is no relationship between fires and bee count.
Alternative Hypothesis: There is a relationship between fires and bee count.
```{r}
# Extract number of rows for before fire data
umpqua_before_fire <- bees_in_buffer_zone(bees_before_fire(9), umpqua_2020)
hood_before_fire <- bees_in_buffer_zone(bees_before_fire(9), hood_2020)
holiday_before_fire <- bees_in_buffer_zone(bees_before_fire(9), holiday_2020)
beachie_before_fire <- bees_in_buffer_zone(bees_before_fire(8), beachie_2020)
freemont_before_fire <- bees_in_buffer_zone(bees_before_fire(8), freemont_2020)
ben_young_before_fire <- bees_in_buffer_zone(bees_before_fire(7), ben_young_2020)
brattain_before_fire <- bees_in_buffer_zone(bees_before_fire(9), brattain_2020)
lionshead_before_fire <- bees_in_buffer_zone(bees_before_fire(8), lionshead_2020)
thielsen_before_fire <- bees_in_buffer_zone(bees_before_fire(9), thielsen_2020)
two_four_before_fire <- bees_in_buffer_zone(bees_before_fire(9), two_four_2020)
whiteriver_before_fire <- bees_in_buffer_zone(bees_before_fire(8), whiteriver_2020)

# Extract number of rows for after fire data
umpqua_after_fire <- bees_in_buffer_zone(bees_after_fire(9), umpqua_2020)
hood_after_fire <- bees_in_buffer_zone(bees_after_fire(9), hood_2020)
holiday_after_fire <- bees_in_buffer_zone(bees_after_fire(9), holiday_2020)
beachie_after_fire <- bees_in_buffer_zone(bees_after_fire(8), beachie_2020)
freemont_after_fire <- bees_in_buffer_zone(bees_after_fire(8), freemont_2020)
ben_young_after_fire <- bees_in_buffer_zone(bees_after_fire(7), ben_young_2020)
brattain_after_fire <- bees_in_buffer_zone(bees_after_fire(9), brattain_2020)
lionshead_after_fire <- bees_in_buffer_zone(bees_after_fire(8), lionshead_2020)
thielsen_after_fire <- bees_in_buffer_zone(bees_after_fire(9), thielsen_2020)
two_four_after_fire <- bees_in_buffer_zone(bees_after_fire(9), two_four_2020)
whiteriver_after_fire <- bees_in_buffer_zone(bees_after_fire(8), whiteriver_2020)
```

```{r}

# Combine before and after data into vectors
before_fire <- c(
  umpqua_before_fire, hood_before_fire, holiday_before_fire, 
  beachie_before_fire, freemont_before_fire, ben_young_before_fire, 
  brattain_before_fire, lionshead_before_fire, thielsen_before_fire, 
  two_four_before_fire, whiteriver_before_fire
)

after_fire <- c(
  umpqua_after_fire, hood_after_fire, holiday_after_fire, 
  beachie_after_fire, freemont_after_fire, ben_young_after_fire, 
  brattain_after_fire, lionshead_after_fire, thielsen_after_fire, 
  two_four_after_fire, whiteriver_after_fire
)

before_fire
after_fire
```

```{r}
# Perform a paired t-test
t_test <- t.test(before_fire, after_fire, paired = TRUE)

# Print the results
cat("Paired T-Test Results:\n")
cat("T-statistic:", t_test$statistic, "\n")
cat("P-value:", t_test$p.value, "\n")
cat("Confidence Interval:", t_test$conf.int, "\n")
cat("Mean difference:", t_test$estimate, "\n")

# Interpretation
if (t_test$p.value < 0.05) {
  cat("Result: Reject the null hypothesis. Bee counts differ significantly before and after fires.\n")
} else {
  cat("Result: Fail to reject the null hypothesis. No significant difference in bee counts before and after fires.\n")
}
```


```{r}
# Calculate p-value for paired t-test
t_test <- t.test(before_fire, after_fire, paired = TRUE)
p_value <- t_test$p.value

# Create a data frame for visualization
visual_data <- data.frame(
  Location = rep(
    c("Umpqua", "Hood", "Holiday", "Beachie", "Freemont", 
      "Ben Young", "Brattain", "Lionshead", "Thielsen", 
      "Two Four", "White River"), 
    times = 2
  ),
  Count = c(before_fire, after_fire),
  Time = rep(c("Before Fire", "After Fire"), each = length(before_fire))
)

# Factor the Time variable to ensure "Before Fire" comes first
visual_data$Time <- factor(visual_data$Time, levels = c("Before Fire", "After Fire"))

# Create the grouped bar plot
ggplot(visual_data, aes(x = Location, y = Count, fill = Time)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7), color = "black") +  # Slight gap between bars
  scale_fill_manual(values = c("Before Fire" = "#1f78b4", "After Fire" = "#e31a1c")) +
  labs(
    title = "Bee Counts Before and After Fires",
    x = "Fire Location",
    y = "Bee Count",
    fill = "Time",
    caption = "Data source: Oregon Bee Atlas"
  ) +
  annotate(
    "text", 
    x = 9, y = max(c(before_fire, after_fire)) * 1.05, 
    label = paste("P-value:", format(p_value, digits = 2)), 
    size = 5, color = "black", fontface = "italic"
  ) +  # Add p-value annotation in the upper-right corner
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 12),
    plot.caption = element_text(size = 10, hjust = 1, face = "italic")
  )


```