---
title: "exploration"
output: html_document
date: "2024-11-20"
---

```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
```

```{r}
# Load 2017 fires
rebel_2017 = rast("data/2017/2017_rebel_sbs/rebel_sbs2.tif")
columbia_2017 = rast("data/2017/columbia river gorge 2017/eaglecreek_sbs.tif")
rogue_2017 = rast("data/2017/rogue river 2017/blanket_creek_sbs.tif")
umpqua_2017 = rast("data/2017/umpqua 2017/brokentooth_sbs.tif")
willamette_2017 = rast("data/2017/willamette 2017/avenue_sbs2.tif")

# Load 2018 fires
rogue_2018 = rast("data/2018/rogue river fire july 2018/hendrix_sbs.tif")
umpqua_2018 = rast("data/2018/umpqua fire july 2018/columbus_sbs.tif")
willamette_2018 = rast("data/2018/willamette national forest fire august 2018/SoilSeverity.tif")
terwilliger_2018 = rast("data/2018/2018_terwilliger_sbs/SoilSeverity.tif")
freemont_2018 = rast("data/2018/freemont-winema 2018/WatsonCreek_SBS.tif")

# Load 2019 fires
blueridge_2019 = rast("data/2019/blueridge malheur 2019/Blue_Ridge_SBS.tif")
cow_2019 = rast("data/2019/cow malheur 2019/204Cow_Mask.tif")
umatilla_2019 = rast("data/2019/umatilla national forest 2019/SoilBurnSeverityFinal.tif")

# Load 2020 fires
beachie_2020 = rast("data/2020/beachie creek willamette 2020/BeachieCreek_SBS_final.tif")
freemont_2020 = rast("data/2020/freemont winema 2020/CRANE_sbs.tif")
holiday_2020 = rast("data/2020/holiday farm willamette 2020/HolidayFarm_SBS_final.tif")
hood_2020 = rast("data/2020/mt hood 2020/Riverside_SBS_final.tif")
umpqua_2020 = rast("data/2020/umpqua 2020/ArchieCreek_SBS_final.tif")
```

```{r}
# Resolutions 2017 fires
res(rebel_2017) # 30 30
res(columbia_2017) # 30 30
res(rogue_2017) # 30 30
res(umpqua_2017) # 30 30 
res(willamette_2017) # 30 30

# Resolutions 2018 fires
res(rogue_2018) # 30 30
res(umpqua_2018) # 20 20
res(willamette_2018) # 30 30
res(terwilliger_2018) # 30 30
res(freemont_2018) # 30 30

# Resolutions 2019 fires
res(blueridge_2019) # 20 20
res(cow_2019) # 20 20
res(umatilla_2019) # 20 20

# Resolutions 2020 fires
res(beachie_2020) # 20 20
res(freemont_2020) # 20 20
res(holiday_2020) # 20 20
res(hood_2020) # 20 20 
res(umpqua_2020) # 20 20
```

```{r}
# Adjust the scale of all fire rasters to be 1-4
adjust_scale <- function(raster) {
    raster[raster > 4] <- NA
}

# Align raster extents to a reference raster
align_extent <- function(raster, reference_raster) {
  extend(raster, ext(reference_raster))
}

# Apply the function for each fire
adjust_scale(rebel_2017)
adjust_scale(columbia_2017)
adjust_scale(rogue_2017)
adjust_scale(umpqua_2017)
adjust_scale(willamette_2017)

adjust_scale(rogue_2018)
adjust_scale(umpqua_2018)
adjust_scale(willamette_2018)
adjust_scale(terwilliger_2018)
adjust_scale(freemont_2018)

adjust_scale(blueridge_2019)
adjust_scale(cow_2019)
adjust_scale(umatilla_2019)

adjust_scale(beachie_2020)
adjust_scale(freemont_2020)
adjust_scale(holiday_2020)
adjust_scale(hood_2020)
adjust_scale(umpqua_2020)
```

```{r}
# Reproject raster data function
reproject_raster <- function(raster, reference_raster) {
  project(raster, crs(reference_raster), res = res(reference_raster))
}

# Convert raster to dataframe function
raster_to_dataframe <- function(raster) {
  as.data.frame(raster, xy = TRUE)
}

# Assign nuneric values function for categorical variables for consistency
transform_soil_burn <- function(df) {
  # Check if 'SoilBurnSe' exists in the dataframe
  if ("SoilBurnSe" %in% colnames(df)) {
    df$SoilBurnSe <- as.character(df$SoilBurnSe)
    df$SoilBurnSe[df$SoilBurnSe == "High"] <- "4"
    df$SoilBurnSe[df$SoilBurnSe == "Moderate"] <- "3"
    df$SoilBurnSe[df$SoilBurnSe == "Low"] <- "2"
    df$SoilBurnSe[df$SoilBurnSe == "Unburned"] <- "1"
    df$SoilBurnSe <- as.factor(df$SoilBurnSe)
  } else{}
  return(df)
}

# Load and transform to a shapefile function
load_and_transform_shapefile <- function(shapefile_path, reference_raster) {
  shapefile <- st_read(shapefile_path)
  st_transform(shapefile, crs(reference_raster))
}

# Generate and save file function
plot_fires_and_bees <- function(willamette_df, rogue_df, umpqua_df, ecoregions, bee_sf, output_file) {
  ggplot() +
    geom_raster(data = willamette_df, aes(x = x, y = y, fill = SoilBurnSe)) +
    geom_raster(data = rogue_df, aes(x = x, y = y, fill = Layer_1)) +
    geom_raster(data = umpqua_df, aes(x = x, y = y, fill = Layer_1)) + 
    geom_sf(data = ecoregions, fill = NA, color = "black") + 
    geom_sf(data = bee_sf, aes(color = Collection.Month), size = 3) +
    scale_fill_brewer(palette = "Spectral", direction = -1) +
    scale_color_viridis_d(name = "Collection Month & Year") +
    labs(
      title = "Map of Longula Bee Species Locations by Collection Month and Year with 3 2018 Fires Plotted",
      x = "Longitude",
      y = "Latitude"
    ) +
    annotation_scale(location = "br", width_hint = 0.2)
  
  ggsave(output_file, width = 8, height = 6, dpi = 300)
}

# Function to process a list of rasters
process_rasters <- function(raster, reference_raster) {
  # Reproject the raster
  raster <- reproject_raster(raster, reference_raster)
    
  # Convert raster to data frame
  raster_df <- raster_to_dataframe(raster)

  return(raster_df)
}
```

```{r}
# Apply the function for each fire
rebel_2017_df <- process_rasters(rebel_2017, willamette_2017)
columbia_2017_df <- process_rasters(columbia_2017, willamette_2017)
rogue_2017_df <- process_rasters(rogue_2017, willamette_2017)
umpqua_2017_df <- process_rasters(umpqua_2017, willamette_2017)
willamette_2017_df <- process_rasters(willamette_2017, willamette_2017)

rogue_2018_df <- process_rasters(rogue_2018, willamette_2017)
umpqua_2018_df <- process_rasters(umpqua_2018, willamette_2017)
willamette_2018_df <- process_rasters(willamette_2018, willamette_2017)
terwilliger_2018_df <- process_rasters(terwilliger_2018, willamette_2017)
freemont_2018_df <- process_rasters(freemont_2018, willamette_2017)

blueridge_2019_df <- process_rasters(blueridge_2019, willamette_2017)
cow_2019_df <- process_rasters(cow_2019, willamette_2017)
umatilla_2019_df <- process_rasters(umatilla_2019, willamette_2017)

beachie_2020_df <- process_rasters(beachie_2020, willamette_2017)
freemont_2020_df <- process_rasters(freemont_2020, willamette_2017)
holiday_2020_df <- process_rasters(holiday_2020, willamette_2017)
hood_2020_df <- process_rasters(hood_2020, willamette_2017)
umpqua_2020_df <- process_rasters(umpqua_2020, willamette_2017)

# Convert the columns to numeric (continuous) if they aren't already
rebel_2017_df$Layer_1 <- as.numeric(rebel_2017_df$Layer_1)
columbia_2017_df$Layer_1 <- as.numeric(columbia_2017_df$Layer_1)
rogue_2017_df$Layer_1 <- as.numeric(rogue_2017_df$Layer_1)
umpqua_2017_df$Layer_1 <- as.numeric(umpqua_2017_df$Layer_1)
willamette_2017_df$Layer_1 <- as.numeric(willamette_2017_df$Layer_1)

rogue_2018_df$Layer_1 <- as.numeric(rogue_2018_df$Layer_1)
umpqua_2018_df$Layer_1 <- as.numeric(umpqua_2018_df$Layer_1)
willamette_2018_df$SoilBurnSe <- as.numeric(willamette_2018_df$SoilBurnSe)
terwilliger_2018_df$SoilBurnSe <- as.numeric(terwilliger_2018_df$SoilBurnSe)
freemont_2018_df$Layer_1 <- as.numeric(freemont_2018_df$Layer_1)

blueridge_2019_df$Layer_1 <- as.numeric(blueridge_2019_df$Layer_1)
cow_2019_df$`204Cow_Mask` <- as.numeric(cow_2019_df$`204Cow_Mask`)
umatilla_2019_df$SoilBurnSeverityFinal <- as.numeric(umatilla_2019_df$SoilBurnSeverityFinal)

beachie_2020_df$Layer_1 <- as.numeric(beachie_2020_df$Layer_1)
freemont_2020_df$Layer_1 <- as.numeric(freemont_2020_df$Layer_1)
holiday_2020_df$Layer_1 <- as.numeric(holiday_2020_df$Layer_1)
hood_2020_df$Layer_1 <- as.numeric(hood_2020_df$Layer_1)
umpqua_2020_df$Layer_1 <- as.numeric(umpqua_2020_df$Layer_1)
```

```{r}
oregon_ecoregions <- st_read("data/OR-ecoregions/Ecoregions_OregonConservationStrategy.shp")
oregon_ecoregions <- st_transform(oregon_ecoregions, crs(willamette_2017))

ggplot() +
  geom_raster(data = rebel_2017_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = columbia_2017_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = rogue_2017_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = umpqua_2017_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = willamette_2017_df, aes(x = x, y = y, fill = Layer_1)) +
  
  geom_raster(data = rogue_2018_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = umpqua_2018_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = willamette_2018_df, aes(x = x, y = y, fill = SoilBurnSe)) +
  geom_raster(data = terwilliger_2018_df, aes(x = x, y = y, fill = SoilBurnSe)) +
  geom_raster(data = freemont_2018_df, aes(x = x, y = y, fill = Layer_1)) +
  
  geom_raster(data = blueridge_2019_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = cow_2019_df, aes(x = x, y = y, fill = `204Cow_Mask`)) +
  geom_raster(data = umatilla_2019_df, aes(x = x, y = y, fill = SoilBurnSeverityFinal)) +
  
  geom_raster(data = beachie_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = freemont_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = holiday_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = hood_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  geom_raster(data = umpqua_2020_df, aes(x = x, y = y, fill = Layer_1)) +
  
  # Add the Oregon ecoregions as a boundary layer
  geom_sf(data = oregon_ecoregions, fill = NA, color = "black") +  
  
  # Customize colors
  scale_fill_gradient(low = "blue", high = "red", name = "Intensity") +  # Set legend title to "Intensity"

  # Add labels and scale
  labs(
    title = "Map of Fire Impacts and Bee Species Locations",
    x = "Longitude",
    y = "Latitude"
  ) +
  
  # Add annotation for scale
  annotation_scale(location = "br", width_hint = 0.2) +
  
  theme_minimal() +
  
  # Customize the theme (optional)
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10))

# Save the plot
ggsave("Oregon_Fires_and_Bees.png", width = 8, height = 6, dpi = 300)
```

### Initial look at the Oregon Bee Atlas data
First, let's read the OBA data and see what columns we have
```{r}
oba_df <- read.csv("data/OBA_2018-2023.csv")
str(oba_df)
```
Now, let's plot all of the bees that are within Oregon on top of a map to see potential areas of interest and Genus distribution.
```{r}
# Step 1: Retrieve Oregon county polygon data
or <- map_data("county", "oregon") %>%  
  select(lon = long, lat, group, id = subregion)

# Convert the Oregon polygon to an sf object (simple feature)
or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326)

# Step 2: Convert your bee data (`oba_df`) to an sf object
## Also going to normalize Genus and Species values
bees_sf <- oba_df %>%
  select(Dec..Lat., Dec..Long., Genus, Species) %>%
  rename(Latitude = Dec..Lat., Longitude = Dec..Long.) %>%
  mutate(
    Genus = tolower(trimws(Genus)), Species = tolower(trimws(Species))
  ) %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)


# Step 3: Crop the bee data to the Oregon bounds
bees_in_oregon <- st_crop(bees_sf, or_sf)

# Step 4: Plot the polygon (oregon), then plot the spatial fetaures (bee locations) on top
ggplot() +
  # Plot Oregon polygons
  geom_polygon(data = or, aes(x = lon, y = lat, group = group), 
               fill = "lightgreen", color = "darkgreen") +
  # Overlay bee locations with coloring by genus
  geom_sf(data = bees_in_oregon, aes(color = Genus)) +
  labs(title = "Bees in Oregon", x = "Longitude", y = "Latitude") +
  theme(
    legend.key.size = unit(0.4, "cm"), # Reduce key size
    legend.text = element_text(size = 4), # Reduce text size
    legend.title = element_text(size = 8) # Reduce title size
  ) +
  guides(color = guide_legend(ncol = 3))

```

### Plot the fires seperately
We want to:
- Plot the bees in the general region the year before the fire
- Then plot the bees in the general region of the fire after the fire year
- Going to look at fires from just 2020

MonthAb contains the month as a number, as expected
Year.1 contains the year as a YYYY number, as expected
```{r}
# Look at the columns of the bee stuff first
oba_df <- oba_df %>%
  rename(
    year = Year.1,
    month = MonthAb,
    full_date = Collection.Date,
    latitude = Dec..Lat.,
    longitude = Dec..Long.,
    plant = Associated.plant
  )

oba_selection <- oba_df %>%
  select(year, month, Genus, Species, plant, latitude, longitude) %>%
  filter(!is.na(Genus), Genus != "")

oba_selection
```
### Plot the bees in the general vecinity of the fires

```{r bees in a region before and after a fire}

plot_bees_in_fire_region <- function(bee_data, fire_sp_obj, fire_name) {
  ### This function will take in bee data (as a dataframe) and a fire raster
  ### and plot the bees found within a 50 km radius of the provided fire
  ### it will also provide a count of the bees within the radius
  ### The goal is to be able to compare the bees in the radius before and after the fire
  
  
  ## Deal with the base map of Oregon
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  
  ## Make sure fire spatial obj matches Oregon map's CRS
  fire_sp_obj <- project(fire_sp_obj, crs(or_sf))
  
  ## Convert bee data to an sf object & make sure CRS matches
  bees_sf <- bee_data %>%
    mutate(
      Genus = tolower(trimws(Genus)),
      Species = tolower(trimws(Species))
    ) %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  
  bees_sf <- st_transform(bees_sf, crs = st_crs(or_sf))
  
  ## Use the fire's bounds to create a buffer (25 km) & capture all the bees within it by cropping
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 15000)
  
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)
  
  ## Total number of bees in the buffer for comparison (before / after fire)
  total_bees <- nrow(bees_in_fire_region)
  
  ## Plot the Oregon polygon and the bee locations in the fire region
  ggplot() +
    geom_polygon(data = or, aes(x = lon, y = lat, group = group),
                 fill = "lightgreen", color = "darkgreen") +
    
    geom_sf(data = fire_buffer, fill = NA, color = "red", linetype = "dashed") +
    
    geom_sf(data = bees_in_fire_region, size = 2, alpha = 0.7) +
    
    labs(
      title = paste("Bees in Fire Region:", fire_name),
      subtitle = paste("Total bees found in the buffer region:", total_bees),
      x = "Longitude", 
      y = "Latitude"
    )
}


```

```{r see results}

## Two years before the fire
bees_before_2020 <- oba_selection %>%
  filter(year %in% c(2018, 2019))

## Two years after the fire
bees_after_2020 <- oba_selection %>%
  filter(year %in% c(2021, 2022))

# Plot the bees before and after 2020 for these 3 fires that were in 2020
plot_bees_in_fire_region(bees_before_2020, umpqua_2020, "Umpqua Fire (before 2020)")
plot_bees_in_fire_region(bees_after_2020, umpqua_2020, "Umpqua Fire (after 2020)")

plot_bees_in_fire_region(bees_before_2020, hood_2020, "Hood Fire (before 2020)")
plot_bees_in_fire_region(bees_after_2020, hood_2020, "Hood Fire (after 2020)")

plot_bees_in_fire_region(bees_before_2020, holiday_2020, "Holiday Fire (before 2020)")
plot_bees_in_fire_region(bees_after_2020, holiday_2020, "Holiday Fire (after 2020)")

plot_bees_in_fire_region(bees_before_2020, beachie_2020, "Beachie Fire (before 2020)")
plot_bees_in_fire_region(bees_after_2020, beachie_2020, "Beachie Fire (after 2020)")

```

### Define the "average severity" for a given fire

```{r average severity}
define_avg_severity <- function(fire_df) {
  ### This function takes all the severity data ("Layer_1" in the raster)
  ### returns the average severity for the fire
  most_severe <- fire_df %>% filter(Layer_1 == 5)
  avg_severity <- sum(most_severe$Layer_1) / nrow(fire_df)
  return(avg_severity)
}


define_avg_severity(umpqua_2020_df)
define_avg_severity(hood_2020_df)
define_avg_severity(holiday_2020_df)
define_avg_severity(beachie_2020_df)
```
## Statistical test to evaluate significance in change
```{r test}
# Function to load and transform the base map of Oregon
load_oregon_map <- function() {
  or <- map_data("county", "oregon") %>%
    select(lon = long, lat = lat, group = group, id = subregion)
  
  or_sf <- st_as_sf(or, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
  return(or_sf)
}

# Function to ensure fire spatial object matches the base map's CRS
project_fire_to_crs <- function(fire_sp_obj, oregon_sf) {
  fire_sp_obj <- project(fire_sp_obj, crs(oregon_sf))
  return(fire_sp_obj)
}

# Function to convert bee data to sf object and transform CRS
convert_bee_data_to_sf <- function(bee_data, oregon_sf) {
  bees_sf <- bee_data %>%
    mutate(
      Genus = tolower(trimws(Genus)),
      Species = tolower(trimws(Species))
    ) %>%
    filter(!is.na(latitude) & !is.na(longitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"), crs = 4326)
  
  bees_sf <- st_transform(bees_sf, crs = st_crs(oregon_sf))
  return(bees_sf)
}

# Function to create a buffer around the fire's bounding box
create_fire_buffer <- function(fire_sp_obj) {
  fire_bbox <- st_as_sfc(st_bbox(fire_sp_obj))
  fire_buffer <- st_buffer(fire_bbox, dist = 25000)  # 50 km buffer
  return(fire_buffer)
}

# Function to calculate bees in the buffer zone
bees_in_buffer_zone <- function(bee_data, fire_sp_obj, oregon_sf) {
  # Load and transform data
  or_sf <- load_oregon_map()
  fire_sp_obj <- project_fire_to_crs(fire_sp_obj, or_sf)
  bees_sf <- convert_bee_data_to_sf(bee_data, or_sf)
  
  # Create fire buffer and calculate intersection with bee data
  fire_buffer <- create_fire_buffer(fire_sp_obj)
  bees_in_fire_region <- st_intersection(bees_sf, fire_buffer)
  
  # Return total number of bees in the fire buffer region
  total_bees <- nrow(bees_in_fire_region)
  return(total_bees)
}

# Function to calculate percent change
calculate_percent_change <- function(before, after) {
  perc_change <- ((after - before) / before) * 100
  return(perc_change)
}

# Usage to calculate percent change in bee populations for 2020
umpqua_before <- bees_in_buffer_zone(bees_before_2020, umpqua_2020, oregon_sf)
umpqua_after <- bees_in_buffer_zone(bees_after_2020, umpqua_2020, oregon_sf)
umpqua_perc_change <- calculate_percent_change(umpqua_before, umpqua_after)

hood_before <- bees_in_buffer_zone(bees_before_2020, hood_2020, oregon_sf)
hood_after <- bees_in_buffer_zone(bees_after_2020, hood_2020, oregon_sf)
hood_perc_change <- calculate_percent_change(hood_before, hood_after)

holiday_before <- bees_in_buffer_zone(bees_before_2020, holiday_2020, oregon_sf)
holiday_after <- bees_in_buffer_zone(bees_after_2020, holiday_2020, oregon_sf)
holiday_perc_change <- calculate_percent_change(holiday_before, holiday_after)

beachie_before <- bees_in_buffer_zone(bees_before_2020, beachie_2020, oregon_sf)
beachie_after <- bees_in_buffer_zone(bees_after_2020, beachie_2020, oregon_sf)
beachie_perc_change <- calculate_percent_change(beachie_before, beachie_after)

# Display all results for the 2020 fires
print("Umpqua average severity and percent change in bees")
define_avg_severity(umpqua_2020_df)
umpqua_perc_change

print("Hood average severity and percent change in bees")
define_avg_severity(hood_2020_df)
hood_perc_change

print("Holiday average severity and percent change in bees")
define_avg_severity(holiday_2020_df)
holiday_perc_change

print("Beachie average severity and percent change in bees")
define_avg_severity(beachie_2020_df)
beachie_perc_change
```
